\section{The Real Numbers}
\label{sec:real-numbers}

\subsection{Defining the Real Numbers}
\label{sec:defining-real-numbers}

Recall that we defined \(\NN\) using the Peano's Axioms, and we may obtain \(\ZZ\) from \(\NN\) by allowing for 'subtraction'.

\begin{definition}[\(\ZZ\)]
    Formally, \(\ZZ\) is the set of equivaleence classes of \(\NN \times \NN\) under the equivalence relation
    \[
        \para{a, b} \sim \para{c, d} \iff a + d = b + c.
    \]
\end{definition}

We may think of \(\para{a, b}\) as \(a - b\). We denote
\[
    0 = \brac{\para{1, 1}}
\]
and
\[
    -a = \brac{\para{1, a + 1}}.
\]

We may similarly define addition and multiplication on \(\times\) using this pair, as
\[
    \para{a, b} + \para{c, d} = \para{a + c, b + d}
\]
and
\[
    \para{a, b} \times \para{c, d} = \para{ac + bd, bc + ad},
\]
and we may indeed check that these operations are well-defined.

Similarly, we may obtain \(\QQ\) from \(\ZZ\) by allowing for 'division'.

\begin{definition}[\(\QQ\)]
    Formally, \(\QQ\) is the set of the equivalence classes of \(\ZZ \times \NN\) under the equivalence relation
    \[
        \para{a, b} \sim \para{c, d} \iff ad = bc.
    \]
\end{definition}

Similarly, notational wise, we write \(\frac{a}{b}\) instead of \(\brac{\para{a, b}}\), and we may similarly define \(+\) and \(\times\) by
\[
    \para{a, b} + \para{c, d} = \para{ad + bc, bd}
\]
and
\[
    \para{a, b} \times \para{c, d} = \para{a \times c, b \times d}.
\]

In addition to this, we may define an \emph{ordering} on \(\QQ\), \(<\), which satisfy the following properties, that
\begin{enumerate}
    \item if \(a, b \in \QQ\), then precisely one of \(a < b, a = b, a > b\) holds;
    \item if \(a < b\) and \(b < c\), then \(a < c\).
\end{enumerate}

This ordering is useful, in the sense that there is a rational number between any two rational numbers. If \(p, q \in \QQ\) with \(p < q\), w e have
\[
    p < \frac{p + q}{2} < q
\]
so \(\QQ\) is \emph{densely ordered}.

However, this does not ensure that all the gaps are filled in.

\begin{proposition}[\(\sqrt{2}\) is irrational]
    There is no rational \(x\) with \(x^2 = 2\).    
\end{proposition}

\begin{proof}[by Number Theory]
    Suppose that \(x^2 = 2\). Note we may assume \(x > 0\) since \(\para{-x}^2 = x^2\). If \(x\) is rational, then
    \[
        x = \frac{a}{b}
    \]
    for some \(a, b \in \NN\). Thus, \(a^2 = 2b^2\).

    But note that the exponent of \(2\) in the prime factorisation on the left-hand side is even, while the exponent of \(2\) in the prime factorisation on the right-hand side is odd. This contradicts with the fundamental theorem of arithemetics.
\end{proof}

\begin{remark}
    This proof reflects from a number theory perspective, that if there is \(x \in \QQ\) with \(x^2 = n\) for some \(n \in \NN\), then \(n\) must itself be a square number.
\end{remark}

\begin{proof}[by Algebra]
    Suppose \(x^2 = 2\) for some \(x = \frac{a}{b}\), with \(a, b \in \NN\). Then, for any \(c, d \in \ZZ\), \(\para{cx + d}\) is of the form \(\frac{e}{b}\) for some \(e \in \ZZ\).

    Then, if \(cx + d > 0\), we have \(cx + d \geq \frac{1}{b}\).

    But \(0 < x - 1 < 1\) as \(1 < x < 2\), if \(n\) is sufficiently large, we have
    \[
        0 < \para{x - 1}^n < \frac{1}{b}.
    \]

    But for any \(n \in \NN\), \(\para{x - 1}^n\) is of the form \(cx + d\), for some \(c, d \in \ZZ\), since \(x^2 = 2\). This leads to a contradiction.
\end{proof}

\begin{remark}
    This proof shows that if the square root of an integer is rational, then it must in fact be a whole number.
\end{remark}

This shows that \(\QQ\) has a gap. But how do we fill in this gap only making reference to \(\QQ\), without mentioning \(\RR\)? (We haven't defined the real numbers yet.)

\(\QQ\) fails another property which is key to constructing the reals. Let \(A\) be the set of positive rationals \(p\), such that \(p^2 < 2\). Formally, let
\[
    A = \set{p \in \QQ}{p^2 < 2}.
\]

We show that \(A\) has no largest number. For any \(p \in A\), consider \(q = p - \frac{p^2 - 2}{p + 2}\).

Since \(p \in A\), \(p^2 - 2 < 0\), so we have \(q > p\). Also,
\[
    q^2 - 2 = \frac{2 \para{p^2 - 2}}{\para{p + 2}^2} < 0,
\]
so \(q \in A\).

Similarly, the set
\[
    \set{q \in \QQ}{q > 0, q^2 > 0}
\]
has no smallest element.

The \textbf{crucial point} is not about the \emph{maximum/minimum} of a set -- it is about a \emph{least upper bound} -- which is the "gap". (And luckily, all the "gaps".)

\begin{definition}[\(\RR\)]
    The \emph{real numbers} \(\RR\) are a set with elements \(0, 1 \in \RR\) where \(0 \neq 1\), equipped with operations \(+\) and \(\times\), and an ordering \(<\) satisfying the following axioms:
    \begin{enumerate}
        \item \(\para{\RR, +, 0}\) is an abelian group: \(+\) is commutative and associative with identity \(0\), and inverses exist under \(+\);
        \item \(\para{\RR\multiplicative, \times, 1}\) is an abelian group: \(\times\) is commutative and associative with identity \(1\), and inverses exist under \(\times\) for non-zero elements;
        \item \(\times\) is distributive over \(+\):
        \[
            a \times \para{b + c} = a \times b + a \times c,
        \]
        and \(\para{\RR, +, \times}\) form a field;
        \item \(\para{\RR, <}\) is a totally ordered field: for all \(a, b\), exactly one of \(a < b\), \(a = b\) and \(a > b\) holds; for all \(a, b, c\), if \(a < b\) and \(b < c\), then \(a < c\);
        \item the ordering preserve arithemetics: for all \(a, b, c\), if \(a < b\), then \(a + c < b + c\), and \(ac < bc\) for \(c > 0\);
        \item given any set \(S\) of reals that is non-empty and bounded above, \(S\) has a \emph{least upper bound}.
    \end{enumerate}

    The final axiom is known as the \emph{least upper bound axiom.}
\end{definition}

\begin{definition}[Least Upper Bound]
    We say a set \(S\) is \emph{bounded above} if there is some \(x \in \RR\), such that \(x \geq y\) for all \(y \in S\). Such an element \(x\) is called an \emph{upper bound of \(S\)}.

    We say \(x\) is \emph{the least upper bound} for \(S\), if \(x\) is an upper bound for \(S\), and any other upper bound \(x'\) satisfies \(x' \geq x\).

    We denote this as \(x = \LUB\para{S}\), \(x = \supremum\para{S}\), or most commonly,
    \[
        x = \sup\para{S}.
    \]
\end{definition}

Note that we defined \(\RR\) -- but its existence is not immediately obvious from the definition. This can be seen from a technique called \emph{Dedkind Cuts}.

\begin{remarks}
    \begin{enumerate}
        \item From the axioms, we can check, for example, that \(0 < 1\). Indeed, if not, then \(1 < 0\). Since \(0 \neq 1\), assume on the countrary that \(1 < 0\), so we have
        \[
            0 = 1 - 1 < 0 - 1 = -1,
        \]
        so
        \[
            0 = 0 \cdot \para{-1} < \para{-1} \para{-1} = 1
        \]
        which gives a contradiction.
        \item We may consider \(\QQ\) as \emph{contained} in \(\RR\), by identifying \(\frac{a}{b} \in \QQ\) with \(a \cdot b\inverse \in \RR\).
        \item \(\QQ\) satisfies all axioms \emph{apart from} the final one -- the least upper bound axiom. For example, the set \(A\) involving squaring to \(2\) above does not have a supremum.
        \item The words non-empty and bounded above are crucial.
        
        If \(S\) is empty, then every \(x \in \RR\) is trivially an upper bound, so there is no least upper bound.

        If \(S\) is not bounded above, then it has no upper bound, and certainly does not have a least one.
    \end{enumerate}
\end{remarks}

We give some examples on the idea of the least upper bound.

\begin{examples}
    \begin{enumerate}
        \item Consider \(S = \set{x \in \RR}{0 \leq x \leq 1} = \ccintv{0}{1}\).
        
        \(2\) is an upper bound for \(S\), since for all \(x \in S\), \(x \leq 2\).

        \(\frac{3}{4}\) is \emph{not} an upper bound for \(S\), since \(\frac{7}{8} \in S\), and \(\frac{7}{8} > \frac{3}{4}\).

        In fact, we have
        \[
            \sup \para{S} = 1,
        \]
        since \(1\) is an upper bound: for all \(x \in S, x \leq 1\); and every other upper bound is \(\geq 1\), since \(1 \in S\).

        \item Consier \(S = \set{x \in \RR}{0 < x < 1} = \oointv{0}{1}\).
        
        Similarly, \(2\) is an upper bound for \(S\), and \(\frac{3}{4}\) is not.

        In fact, we have
        \[
            \sup \para{S} = 1,
        \]
        since \(1\) is an upper bound: for all \(x \in S, x < 1\) so \(x \leq 1\); and no opper bound \(c\) satisfies \(c < 1\).

        Indeed, \(c\) is certainly greater than \(0\), and furthermore \(c \geq \frac{1}{2}\) since \(\frac{1}{2} \in S\).

        So if \(c < 1\), then \(0 < c < 1\), so \(\frac{c + 1}{2} \in S\), but we have \(\frac{c + 1}{2} > c\).
    \end{enumerate}
\end{examples}

\begin{remark}
    If \(S\) have a greatest element, then
    \[
        \sup \para{S} = \max \para{S} \in S.
    \]

    However, as the example shows, \(\sup\para{S}\) can exist when \(\max\para{S}\) does not: in this case, \(\sup\para{S} \notin S\).
\end{remark}

Now we want to find the least upper bound of
\[
    S = \brce{0, \frac{1}{2}, \frac{2}{3}, \frac{3}{4}, \ldots} = \set{1 - \frac{1}{n}}{n \in \NN}.
\]

Clearly, \(1\) is an upper bound of \(S\). But is there an upper bound less than 1?

\begin{proposition}[Axiom of Archimedes]
    \(\NN\) is not bounded above in \(\RR\).
\end{proposition}

\begin{proof}
    Suppose on the contrary, that \(\NN\) is bounded above, then it has a least upper bound
    \[
        c = \sup \para{\NN} \in \RR.
    \]

    By definition, \(\para{c - 1}\) is not an upper bound for \(\NN\), so there is some \(n \in \NN\), such that \(n > c - 1\).

    Then, \(n + 1 \in \NN\), with \(n + 1 > c\), contradicting that \(c\) is an upper bound for \(\NN\).
\end{proof}

\begin{corollary}[Reciprocal of Natural Numbers can be Arbitrarily Small]
    For any real number \(t > 0\), there exists \(n \in \NN\) such that \(\frac{1}{n} < t\).
\end{corollary}

\begin{proof}
    Given \(t > 0\), by Axiom of Archimedes, there is \(n \in \NN\) where \(n > \frac{1}{t}\).

    Hence, \(\frac{1}{n} < t\) as desired.
\end{proof}

\begin{definition}[Greatest Lower Bound]
    A set \(S\) is said to be \emph{bounded below}, if there exists \(x\), such that for all \(y \in S\), we have \(x \leq y\). Such an \(x\) is called a \emph{lower bound} for \(S\).

    We say \(x\) is the \emph{greatest lower bound} for \(S\), if \(x\) is a lower bound for \(S\), and any other lower bound \(x'\) satisfies \(x' \leq x\).

    We denote this as \(x = \GLB \para{S}\), \(x = \infimum\para{S}\), or most commonly,
    \[
        x = \inf \para{S}.
    \]
\end{definition}

Note that if \(S\) is empty and bounded below, then
\[
    -S = \set{-y}{y \in S}
\]
is non-empty and bounded above, so it has a least upper bound
\[
    \sup \para{-S} = c,
\]
and hence \(-c\) is the greatest lower bound.

The previous corollary immediately implies that
\[
    \inf \para{\set{\frac{1}{n}}{n \in \NN}} = 0.
\]

The previous proposition and corollary can be interpreted as that there are no 'infinitely large' or 'infinitely small' numbers in \(\RR\).

Now, we return to a previous example.

\begin{example}
    We want to find the least upper bound of
    \[
        S = \brce{0, \frac{1}{2}, \frac{2}{3}, \frac{3}{4}, \ldots} = \set{1 - \frac{1}{n}}{n \in \NN}.
    \]

    We claim \(\sup \para{S} = 1\), and suppose on the countrary that \(c < 1\) is an upper bound for \(S\). Then \(1 - \frac{1}{n} \leq c\), for all \(n \in \NN\), so
    \[
        0 < 1 - c \leq \frac{1}{n}
    \]
    for all \(n \in \NN\), contradicting with the previous corollary.
\end{example}

Now we show that the gaps we found in \(\QQ\) beforehand is filled by the least upper bound axiom.

\begin{theorem}[Square Roots Exist in \(\RR\)]
    There exists \(x \in \RR\) with \(x^2 = 2\).
\end{theorem}

\begin{proof}
    Let
    \[
        S = \set{x \in \RR}{x^2 < 2}.
    \]

    First we note that \(S\) is non-empty since \(1 \in S\), and it is bounded above by \(2\)

    Hence, \(S\) has a supremum, and let \(c = \sup S\). We observe \(1 < c < 2\), ad we claim that \(c^2 = 2\).

    Suppose that \(c^2 < 2\). For \(0 < t < 1\), we have
    \begin{align*}
        \para{c + t}^2 &= c^2 + 2ct + t^2 \\
        &< c^2 + 4t + t \\
        &= c^2 + 5t \\
        &< 2,
    \end{align*}
    for sufficiently small \(t\), say \(t < \frac{2 - c^2}{5}\).

    But this contradicts with the assumption that \(c\) is \emph{an upper bound} for \(S\), since \(c + t \in S\) and \(c + t > c\).

    Now, suppose that \(c^2 > 2\). For \(0 < t < 1\), we have
    \begin{align*}
        \para{c - t}^2 &= c^2 - 2ct + t^2 \\
        &> c^2 - 4t + 0 \\
        &> 2,
    \end{align*}
    for sufficiently small \(t\), say \(t < \frac{c^2 - 2}{4}\).

    This contradicts with the assumption that \(c\) is the \emph{least} upper bound of \(S\), since \(c - t < c\) and \(c - t\) is an upper bound for \(S\).

    Hence, \(c^2 = 2\).
\end{proof}

\begin{remark}
    A similar proof shows that \(\sqrt[n]{x}\) exists for all \(n \in \NN\), for all \(x \in \RR\) where \(x > 0\):
    \[
        \forall n \in \NN, \forall x \in \oointv{0}{\plusinfty}, \exists y \in \RR: y^n = x.
    \]
\end{remark}

\begin{definition}[Irrational Numbers]
    A real that is not rational is \emph{irrational}.
\end{definition}

\begin{examples}[Examples of Irrationals]
    \begin{enumerate}
        \item \(\sqrt{2}, \sqrt{3}, \sqrt{5}, \sqrt{6}, \ldots\) are irrational.
        \item If \(2 + 3 \sqrt{5} = \frac{a}{b}\) for some \(a, b \in \NN\), then
        \[
            \sqrt{5} = \frac{a - 2b}{3b} \in \QQ
        \]
        but \(\sqrt{5}\) is irrational. So \(2 + 3 \sqrt{5}\) is irrational.
    \end{enumerate}
\end{examples}

\begin{proposition}[\(\QQ\) is dense in \(\RR\)]
    The rationals are \emph{dense} in the reals. Formally,
    \[
        \forall a, b \in \RR, a < b, \exists c \in \QQ: a < c < b.
    \]
\end{proposition}

\begin{proof}
    We may assume \(a \geq 0\). By corollary of the axiom of achimedes, we have some \(n \in \NN\) such that \(\frac{1}{n} < b - a\).

    Let
    \[
        T = \set{k \in \NN}{\frac{k}{n} \geq b}.
    \]

    By axiom of archimedes, there is some \(N \in \NN\) such that \(N > b\).
    
    Hence, we have
    \[
        \frac{Nn}{n} \geq b,
    \]
    and hence \(Nn \in T\), so \(T \neq \emptyset\).

    By the well-ordering principle, \(T\) has some least element \(m\). Now, let \(c = \frac{m - 1}{n} \in \QQ\).

    Since \(m - 1 \notin T\), we have \(c < b\).

    Now, if \(c \leq a\), then
    \[
        \frac{m}{n} = c + \frac{1}{n} < a + \para{b - a} = b,
    \]
    contradicts with \(m \in T\).

    Therefore, \(c > a\), and hence \(a < c < b\) with \(c \in \QQ\), as desired.
\end{proof}

\begin{proposition}[\(\RR \setminus \QQ\) is dense in \(\RR\)]
    The irrationals are \emph{dense} in the reals. Formally,
    \[
        \forall a, b \in \RR, a < b, \exists c \in \RR \setminus \QQ: a < c < b.
    \]
\end{proposition}

\begin{proof}
    Indeed, take \(c \in \QQ\) such that
    \[
        a \sqrt{2} < c < b \sqrt{2}
    \]
    then
    \[
        a < \frac{c}{\sqrt{2}} < b.
    \]
\end{proof}

\subsection{Sequences}
\label{sec:sequences}

\begin{definition}[Sequence]
    A \emph{sequence} is an enumerated collection of objects in which repetitions are allowed and order matters.

    We write the sequence as
    \[
        a_1, a_2, a_3, \ldots
    \]
    or
    \[
        \para{a_n}_{n = 1}^{\infty}.
    \]
\end{definition}

We aim to answer the question of what it means for a sequence \(a_1, a_2, a_3, \ldots\) to tend to a limit \(l\).

First, it is \emph{not} enough on its own that the terms get 'arbitrarily close' to \(l\), in the sense that
\[
    \forall \epsilon > 0, \exists n \in \NN: l - \epsilon < a_n < l + \epsilon.
\]

For example, we do not want the sequence
\[
    0, 10, \frac{1}{2}, 10, \frac{2}{3}, 10, \frac{3}{4}, 10, \ldots
\]
to tend to \(1\).

We want the sequence to get to, and stay within \(\epsilon\) of \(l\).

\begin{definition}[Limit of a Sequence]
    We say the sequence \(a_1, a_2, a_3, \ldots\) tends to a \emph{limit} \(l \in \RR\), as \(n\) tends to infinity, if
    \[
        \forall \epsilon > 0, \exists N \in \NN, \forall n \geq N: l - \epsilon < a_n < l + \epsilon.
    \]

    More compactly, using the absolute value
    \[
        \abs{x} = \begin{cases}
            x, & x \geq 0, \\
            -x, & x < 0,
        \end{cases}
    \]
    the definition is equivalent to
    \[
        \forall \epsilon > 0, \exists N \in \NN, \forall n \geq N: \abs{a_n - l} < \epsilon.
    \]
\end{definition}

We may think of the absolute value \(\abs{a - b}\) as the \emph{distance} between \(a\) and \(b\) on the real number line, for example,
\[
    \abs{2 - 9} = \abs{9 - 2} = 7.
\]

\begin{proposition}[Triangle Inequality]
    We have that for all \(a, b, c \in \RR\), that
    \[
        \abs{a - c} \leq \abs{a - b} + \abs{b - c}.
    \]
\end{proposition}

\begin{notation}
    When \(\para{a_n}_{n = 1}^{\infty}\) tends to \(l\) as \(n\) tends to infinity, we may write
    \[
        a_n \to l \text{ as } n \to \infty,
    \]
    or
    \[
        \lim_{n \to \infty} a_n = l,
    \]
    or
    \[
        \para{a_n}_{n = 1}^{\infty} \text{ converges to } l.
    \]

    If there is a limit \(l\), but it is not specified, we may simply say the sequence converges; if such limit does not converge to any \(l \in \RR\), we say it \emph{diverges}.
\end{notation}

\begin{examples}
    \begin{enumerate}
        \item Consider the sequence \(0, \frac{1}{2}, \frac{2}{3}, \frac{3}{4}, \frac{4}{5}, \ldots\), i.e., \(a_n = 1 - \frac{1}{n}\).
        
        Given \(\epsilon > 0\), choose \(N > \frac{1}{\epsilon}\) (by axiom of archimedes). If \(n \geq N\), then
        \[
            \abs{a_n - 1} = \abs{1 - \frac{1}{n} - 1} = \frac{1}{n} \leq \frac{1}{N} < \epsilon.
        \]

        Hence, \(a_n \to 1\) as \(n \to \infty\).

        \item Consider the sequence \(0, \frac{1}{2}, 0, \frac{1}{4}, 0, \ldots\), i.e.,
        \[
            a_n = \begin{cases}
                \frac{1}{n}, & n \text{ is even},\\
                0, & n \text{ is odd}.
            \end{cases}
        \]

        Given \(\epsilon > 0\), pick \(N > \frac{1}{\epsilon}\). If \(n \geq N\), then
        \[
            \abs{a_n - 0} \leq \frac{1}{n} \leq \frac{1}{N} < \epsilon.
        \]

        Hence, \(a_n \to 0\) as \(n \to \infty\).

        \item Consider the sequence \(a_n = 1 - \frac{1}{2^n}\). Note that it can also be thought of as a \emph{series}
        \[
            \frac{1}{2}, \frac{1}{2} + \frac{1}{4}, \frac{1}{2} + \frac{1}{4} + \frac{1}{8}, \ldots.  
        \]

        Given \(\epsilon . 0\), choose \(N > \frac{1}{\epsilon}\). If \(n \geq N\), then we have
        \[
            \abs{a_n - 1} = \frac{1}{2^n} \leq \frac{1}{n} \leq \frac{1}{N} < \epsilon.
        \]

        Hence, \(a_n \to 1\) as \(n \to \infty\).

        \item Consider the sequence \(a_n = \para{-1}^n\). We aim to show that it does not tend to \emph{any} real number \(l \in \RR\), i.e.,
        \[
            \forall l \in \RR, \exists \epsilon > 0, \forall N \in \NN, \exists n \geq N: \abs{a_n - l} \geq \epsilon.
        \]

        We first show that \(a_n \notto 0\). Let \(\epsilon = 1\), and observe that
        \[
            \abs{a_n - 0} = 1
        \]
        for all \(n \in \NN\).

        In fact, we can show that \(a_n \notto l\) for any \(l \in \RR\). Suppose on the contrary that \(a_n \to l\) as \(n \to \infty\), for some \(l \in \RR\).

        Let \(\epsilon = 1\), and hence
        \[
            \exists N \in \NN, \forall n \geq N: \abs{a_n - l} < 1.
        \]

        In particular, we have
        \[
            \abs{1 - l} < 1, \abs{-1 - l} < 1,
        \]
        and hence by the triangular inequaltiy,
        \begin{align*}
            2 &> \abs{1 - l} + \abs{-1 - l} \\
            &\geq \abs{\para{1 - l} - \para{-1 - l}} \\
            &= \abs{2} \\
            &= 2.
        \end{align*}

        Therefore, \(\para{\para{-1}^n}_{n = 1}^{\infty}\) is divergent. However, it does not "go off to infinity" in that sense.
    \end{enumerate}
\end{examples}

So far, we implicitly assumed that if a limit exists, then it is unique.

\begin{proposition}[Limits are Unique]
    If \(a_n \to l\) and \(a_n \to k\) as \(n \to \infty\), then \(l = k\).
\end{proposition}

\begin{proof}
    Suppose \(l \neq k\), then we choose \(\epsilon = \frac{1}{2} \abs{l - k} > 0\).

    Then there exists \(N \in \NN\), such that
    \[
        \forall n \geq N: \abs{a_n - l} < \epsilon,
    \]
    and there exists \(M \in \NN\), such that
    \[
        \forall n \geq M: \abs{a_n - k} < \epsilon.
    \]

    Hence,
    \begin{align*}
        2\epsilon &> \abs{a_n - l} + \abs{a_n - k} \\
        &\geq \abs{\para{a_n - l} - \para{a_n - k}} \\
        &= \abs{l - k} \\
        &= 2\epsilon,
    \end{align*}
    which gives a contradiction.
\end{proof}

However, we would like to study more generally on the sequences that converge. An important technique is putting \emph{bounds} on the sequence.

\begin{definition}[Bounded Sequence]
    A sequence \(\para{a_n}_{n = 1}^{\infty}\) is \emph{bounded above}, if there exists a real number \(U\), such that \(a_n < U\) for all \(n \in \NN\).

    Similarly, it is \emph{bounded below}, if there exists a real number \(L\), such that \(a_n > L\) for all \(n \in \NN\).

    It is \emph{bounded} if it is both bounded above and below. Alternatively, we may define it to be \emph{bounded} if there exists a real number \(B\), such that \(\abs{a_n} < B\) for all \(n \in \NN\).
\end{definition}

We see the two definitions of bounded are equivalent. Indeed, if \(\abs{a_n} < B\) for some \(B \in \RR\), then let \(U = B\) and \(L = -B\), then it is both bounded above and below. If such \(U, L \in \RR\) exists, then let \(B = \max\brce{\abs{U}, \abs{L}}\).

\begin{proposition}[Convergent Sequences are Bounded]
    A convergent sequence is bounded.
\end{proposition}

\begin{proof}
    Let \(a_n \to l\) as \(n \to \infty\). Then let \(N \in \NN\), such that
    \[
        \forall n \geq N: \abs{a_n - l} < 1.
    \]

    Hence,
    \[
        \abs{a_n} \leq \max\brce{\abs{a_1}, \abs{a_2}, \ldots, \abs{a_{N - 1}}, \abs{l} + 1}
    \]
    for all \(n \in \NN\).
\end{proof}

\begin{definition}[Monotonic]
    A sequnece \(\para{a_n}_{n = 1}^{\infty}\) is \emph{increasing} if \(a_{n + 1} \geq a_n\) for all \(n \in \NN\).

    Similarly, it is \emph{decreasing} if \(a_{n + 1} \leq a_n\) for all \(n \in \NN\).

    It is \emph{monotonic} if it is increasing or if it is decreasing.
\end{definition}

This leads to the following theorem.

\begin{theorem}[Monotone Convergence Theorem]
    Every bounded monotonic sequence of real numbers converges.
\end{theorem}

\begin{proof}
    Suppose \(\para{a_n}_{n = 1}^{\infty}\) is increasing and bounded above. We want to show it converges to its supremum.

    Then the set
    \[
        S = \set{a_n}{n \in \NN}
    \]
    is non-empty and bounded above, so let
    \[
        l = \sup S
    \]

    Let \(\epsilon > 0\) be arbitrary. \(l - \epsilon\) is \emph{not} an upper bound for \(S\), and hence there is some \(N \in \NN\) with
    \[
        a_N > l - \epsilon.
    \]

    Hence, for all \(n \geq N\), we have
    \[
        l - \epsilon < a_n \leq l,
    \]
    since \(\para{a_n}\) is monotonic.

    Hence,
    \[
        \forall n \geq N: \abs{a_n - l} < \epsilon
    \]
    so \(a_n \to l\).

    The decreasing and bounded below case is similar.
\end{proof}

\begin{remarks}
    \begin{enumerate}
        \item For an increasing sequence to converge, we only need to know that it is bounded above. The sequence converges to its supremum.
        \item The sequence \(\para{a_n}_{n = 1}^{\infty}\) with \(a_n = n\) is increasing and not bounded above -- and it diverges.
        \item Monotone convergence theorem is in fact equivalent to the least upper bound axiom -- meaning that if we substitude the latter with the former, we get an equivalent formation of the reals.
        \item Every sequence must have a monotonic subsequence.
    \end{enumerate}
\end{remarks}

\begin{proposition}[Limits Preserve Weak Inequalities]
    If \(a_n \leq d\) for all \(n \in \NN\), and \(a_n \to c\) as \(n \to \infty\), then \(c \leq d\).
\end{proposition}

\begin{proof}
    Suppose that \(c > d\). Let \(\epsilon = c - d > 0\).

    Then by definition of a limit,
    \[
        \exists N \in \NN, \forall n \geq N: \abs{a_n - c} < \epsilon.
    \]

    However, if \(c > d\),
    \[
        \abs{a_n - c} < \epsilon \implies a_n > d.
    \]

    This is a contradiction.
\end{proof}

\begin{remark}
    Limits do not necessarily preserve \emph{strong} inequalities, meaning that if we change the condition \(a_n \leq d\) to \(a_n < d\), then we \text{do not} necessarily have \(\lim a_n < d\) if such limit exists.

    As a counterexample, in the previous case, the sequence
    \[
        \frac{1}{2}, \frac{3}{4}, \frac{7}{8}, \ldots
    \]
    have all terms \(< 1\), but the limit is \(1\).
\end{remark}

We can do limits in addition, if the limits of the separate sequences being summed exists.

\begin{proposition}[Limtis Preserve Sums]
    If \(a_n \to c\) as \(n \to \infty\), and \(b_n \to d\) as \(n \to \infty\), then \(a_n + b_n \to c + d\) as \(n \to \infty\).
\end{proposition}

\begin{proof}
    Let \(\epsilon > 0\) be arbitrary, then
    \[
        \exists N \in \NN, \forall n \geq N: \abs{a_n - c} < \frac{\epsilon}{2},
    \]
    and
    \[
        \exists N \in \NN, \forall n \geq N: \abs{b_n - d} < \frac{\epsilon}{2}.
    \]

    Let \(N' = \max\brce{N, M}\), then for all \(n \geq N'\), we have
    \begin{align*}
        \abs{\para{a_n + b_n} - \para{c + d}} &= \abs{\para{a_n - c} + \para{b_n - d}} \\
        &\leq \abs{a_n - c} + \abs{b_n - d} \\
        &< \frac{\epsilon}{2} + \frac{\epsilon}{2} \\
        &= \epsilon.
    \end{align*}

    This finishes the proof.
\end{proof}

\subsection{Series}
\label{sec:series}

In the reals, the sum of \(2\) numbers is defined, so by induction, finite sums are defined.

We may say things about infinite sums like
\[
    1 - \frac{1}{2} + \frac{1}{3} - \frac{1}{4} + \frac{1}{5} - \cdots = \log 2.
\]

We aim to formalise the study of infinite sums in this section.

\begin{definition}[Partial Sums and Infinite Series]
    Let \(\para{a_n}_{n = 1}^{\infty}\) be a sequence, then
    \[
        s_k = \sum_{n = 1}^{k} a_n
    \]
    is the \(k\)-th \emph{partial sum} of the \emph{series} whose \(n\)-th term is \(a_n\).

    We write
    \[
        \sum_{n = 1}^{\infty} a_n = \lim_{k \to \infty} s_k
    \]
    if the limit exists.
\end{definition}

\begin{examples}[Examples of Series]
    \begin{enumerate}
        \item The series whose \(n\)-th term is \(a_n = r^n\) for some constant \(\abs{r} < 1\) is known as the \emph{geometric series}.
        
        We have
        \begin{align*}
            s_k &= r + r^2 + \cdots + r^k \\
            &= r \cdot \frac{1 - r^k}{1 - r} \\
            & \to r \cdot \frac{1}{1 - r}
        \end{align*}
        as \(k \to \infty\), since \(r^k = 0\).

        Hence,
        \[
            \sum_{n = 1}^{\infty} r^n = \frac{r}{1 - r}.
        \]

        \item The series whose \(n\)-th term is \(a_n = \frac{1}{n}\) is known as the \emph{harmonic series}.
        
        We have
        \begin{align*}
            s_{2^k} &= 1 + \frac{1}{2} + \underbrace{\frac{1}{3} + \frac{1}{4}}_{\text{each} \geq \frac{1}{4}} + \underbrace{\frac{1}{5} + \frac{1}{6} + \frac{1}{7} + \frac{1}{8}}_{\text{each} \geq \frac{1}{8}} + \cdots + \frac{1}{2^k} \\
            &\geq 1 + \frac{1}{2} + 2 \cdot \frac{1}{4} + 4 \cdot \frac{1}{8} + \cdots + \frac{1}{2^k}.
        \end{align*}

        In general, we have
        \[
            \frac{1}{2^m + 1} + \frac{1}{2^m + 2} + \cdots + \frac{1}{2^{m + 1}} \geq 2^m \cdot \frac{1}{2^{m + 1}} = \frac{1}{2}.
        \]

        Hence,
        \[
            s_{2^k} \geq 1 + \frac{k}{2},
        \]
        so the partial sums are increasing and unbounded, and hence the infinite series
        \[
            \sum_{n = 1}^{\infty} \frac{1}{n}
        \]
        diverges.

        \item The series whose \(n\)-th term is \(a_n = \frac{1}{n^2}\) satisfies
        \[
            s_{2^n - 1} = 1 + \underbrace{\frac{1}{2^2} + \frac{1}{3^2}}_{\leq 2 \cdot \frac{1}{2^2}} + \underbrace{\frac{1}{4^2} + \frac{1}{5^2} + \frac{1}{6^2} + \frac{1}{7^2}}_{\leq 4 \cdot \frac{1}{4^2}} + \cdots + \frac{1}{\para{2^n - 1}^2}
        \]
        and in general, we have
        \[
            \frac{1}{\para{2^m}^2} + \frac{1}{\para{2^m + 1}^2} + \cdots + \frac{1}{\para{2^{m + 1} - 1}^2} \leq 2^m \cdot \frac{1}{\para{2^m}^2} = \frac{1}{2^m}.
        \]

        Hence, we have
        \[
            s_{2^k - 1} \leq 1 + \frac{1}{2} + \frac{1}{4} + \cdots + \frac{1}{2^{n - 1}} < 2,
        \]
        and by monotone convergence theorem, we have
        \[
            \sum_{n = 1}^{\infty} \frac{1}{n^2}
        \]
        converges sice its partial sums are increasing and bounded above.

        In fact, we have
        \[
            \sum_{n = 1}^{\infty} \frac{1}{n^2} = \frac{\pi^2}{6}.
        \]
    \end{enumerate}
\end{examples}

From the previous examples, we see that series whose terms are positive are of particular interest, since such series are increasing, and to study their convergence, we just have to study whether they are bounded above or not.

We next study a particular type of sequences in the reals -- the \textbf{decimal expansions}.

\begin{proposition}[Decimal Expansions Converge]
    Let \(\para{d_n}_{n = 1}^{\infty}\) be a sequence with terms \(d_n \in \brce{0, 1, 2, \ldots, 9}\). Then
    \[
        \sum_{n = 1}^{\infty} \frac{d_n}{10^n}
    \]
    converges to some limit \(x\) where \(0 \leq x \leq 1\).
\end{proposition}

\begin{proof}
    The partial sums
    \[
        s_m = \sum_{n = 1}^{m} \frac{d_n}{10^n}
    \]
    are increasing, and they are bounded below (trivially) by \(0\) and above by
    \[
        \sum_{n = 1}^{\infty} \frac{9}{10^n} = \frac{9}{10} \cdot \frac{1}{1 - \frac{1}{10}} = 1.
    \]
\end{proof}

\begin{definition}[Decimal Expansion]
    We say that \(0.d_1 d_2 d_3 \ldots\) is a \emph{decimal expansion} of \(x\).
\end{definition}

The next question to ask, is if the decimal expansions are correspondent to the reals in \(\ccintv{0}{1}\) in some way.

\begin{proposition}[Reals in \(\ccintv{0}{1}\) have Decimal Expansions]
    Any real \(x\) with \(0 \leq x \leq 1\) have a decimal expansion.
\end{proposition}

\begin{proof}
    This is trivially true for \(x = 1\) by \(0.999\ldots\). We consider \(0 \leq x < 1\). We conduct the follwoing algorithm to construct such decimal expansion.

    We pick \(d_1 \in \ZZ\) be maximal such that
    \[
        0 \leq \frac{d_1}{10} \leq x < 1,
    \]
    and hence we have \(0 \leq d_1 \leq 9\), since \(0 \leq x < 1\).

    Furthermore, we have
    \[
        0 \leq x - \frac{d_1}{10} < \frac{1}{10}
    \]
    since \(d_1\) is maximal.

    Then, pick \(d_2 \in \ZZ\) be maximal such that
    \[
        0 \leq \frac{d_2}{100} \leq x - \frac{d_1}{10} < \frac{1}{10},
    \]
    and similarly, \(0 \leq d_2 \leq 9\), and we have
    \[
        0 \leq x - \frac{d_1}{10} - \frac{d_2}{100} < \frac{1}{100}
    \]
    since \(d_2\) is maximal.

    We continue this process indefinitely, picking \(d_n \in \ZZ\) maximal, such that
    \[
        \frac{d_n}{10^n} \leq x - \sum_{j = 1}^{n - 1} \frac{d_j}{10^j}
    \]
    and so
    \[
        0 \leq x - \sum_{j = 1}^{n} \frac{d_j}{10^j} < \frac{1}{10^n}.
    \]

    Since \(\frac{10^n} \to 0\) as \(n \to \infty\), we have
    \[
        x - \sum_{j = 1}^{n} \frac{d_j}{10^j} \to 0,
    \]
    and hence
    \[
        x = \sum_{j = 1}^{\infty} \frac{d_j}{10^j} = 0.d_1 d_2 d_3 \ldots.
    \]
\end{proof}

However, decimal expansions do not have to be unique: a straightforward example is
\[
    1.000\ldots = 0.999\ldots
\]
and anything similar terminating in this. Indeed, these are the only ones.

\begin{proposition}[Decimal Expansions are not Unique precisely for Finite Decimals]
    If a number has two decimal expansions
    \[
        0.a_1 a_2 a_3 \ldots = 0.b_1 b_2 b_3 \ldots
    \]
    that are not identical, then there is some \(k\) such that
    \begin{itemize}
        \item \(a_j = b_j\) for \(j < k\);
        \item \(a_j = b_j - 1\) for \(j = k\); and
        \item \(a_j = 9, b_j = 0\) for \(j > k\);
    \end{itemize}
    or reversed (i.e. swapping the \(a\) and \(b\) above).
\end{proposition}

\begin{proof}
    We suppose \(k\) is maximal such that \(a_j = b_j\) for all \(j < k\), i.e. the expansion is the same precisely up to but not including this digit.

    Without loss of generality, we say \(a_k < b_k\) for the first non-identical digit.

    By bounding, we have
    \begin{align*}
        \sum_{j = k + 1}^{\infty} \frac{a_j}{10^j} &\leq \sum_{j = k + 1}^{\infty} \frac{9}{10^j} \\
        &= \frac{9}{10^{k + 1}} \cdot \frac{1}{1 - \frac{1}{10}} \\
        &= \frac{1}{10^k},
    \end{align*}
    so we must have \(b_k = a_k + 1\).

    (If \(b_k > a_k + 1\), then \(0.b_1 b_2 b_3 \ldots - 0.a_1 a_2 a_3 \ldots \geq \underbrace{2\cdot 10^{-k}}_{k\text{th place}} - \underbrace{10^{-k}}_{\text{tail}} > 0\).)

    For the equal sign to take place, for \(j > k\), we must have
    \[
        a_j = 9, b_j = 0.
    \]
\end{proof}

This case is to be dealt carefully when dealing with reals by considering their decimal expansions.

\begin{definition}[Periodic Decimals]
    A decimal expansion is \emph{periodic} if after a finite number of terms, it repeats in blocks of a fixed length. Equivalently, a decimal expansion \(0.d_1 d_2 d_3 \ldots\) is periodic if
    \[
        \exists l, k \in \NN, \forall n > l: d_{n + k} = d_n.
    \]
\end{definition}

\begin{proposition}[Periodic Decimals are Rational]
    A periodic decimal is rational.
\end{proposition}

\begin{example}[Converting from Decimals to Fractions]
    Consider \(x = 0.\underbrace{7832}_{l}\underbrace{147}\underbrace{147}_{k = 3}\underbrace{147}\).

    We have
    \begin{align*}
        10^4 \cdot x - 7832 &= 0.147 147 147 \ldots \\
        &= 147 \cdot \sum_{j = 1}^{\infty} \frac{1}{10^{3j}} \\
        &= 147 \cdot \frac{1}{10^3} \cdot \frac{1}{1 - \frac{1}{10^3}} \\
        &\in \QQ
    \end{align*}
    so \(x \in \QQ\).
\end{example}

\begin{proposition}[Rationals have Periodic Decimal Expansion]
    A rational has a periodic decimal expansion.
\end{proposition}

\begin{proof}
    Let \(x \in \QQ\) be
    \[
        x = \frac{p}{2^a 5^b q}
    \]
    for \(a, b, p, q \in \ZZ, a, b \geq 0, q > 0\) and \(\gcd\para{q, 10} = 1\).

    Then, we have
    \[
        10^{\max\brce{a, b}} x = \frac{t}{q} = n + \frac{c}{q}
    \]
    for some \(n, c \in \ZZ\), satisfying \(0 \leq c < q\).

    By Fermat--Euler Theorem, since \(\gcd\para{q, 10} = 1\), we have
    \[
        10^{\totient\para{q}} - 1 = kq
    \]
    for some \(k \in \NN\).

    Hence,
    \[
        \frac{c}{q} = \frac{kc}{kq} = \frac{kc}{10^{\totient\para{q}} - 1} = kc \cdot \sum_{j = 1}^{\infty} \frac{1}{\para{10^{\totient\para{q}}}^j},
    \]
    and that \(0 \leq kc < kq\) tells us that \(kc\) is (at most) a \(\totient\para{q}\)-digit number
    \[
        d_1 d_2 \cdots d_{\totient\para{q}},
    \]
    potentially with leading zeroes.

    This therefore gives
    \[
        \frac{c}{q} = 0.d_1 d_2 \cdots d_{\totient\para{q}} d_1 d_2 \cdots d_{\totient\para{q}} \cdots
    \]
    so \(x\) is periodic.
\end{proof}